Compositional Certified Resource Bounds\ifshort{\vspace{-.15cm}}{}

### 2014-07-07 paper/main.tex
Joshua's remakrs for section 2

The idea that drives the design of our framework is amortized
analysis~\cite{Tarjan-amort}.  Assume that a program $S$ executes on a
starting state $\state$ and consumes $n$ resource units of some
quantity we are interested in.  We denote that by writing $(S, \state)
\Downarrow_n \state'$ where $\state'$ is the program state after the
execution.  The basic idea of amortized analysis is to define a
\emph{potential function} $\Phi$ that maps program states to non-negative
numbers and to show that $\Phi(\state) \geq n$ if $\state$ is a
program state such that $(S, \state) \Downarrow_n \state'$.

@@

\caption{Example derivation using amortized reasoning.  We write
  $\#_1(a)$ for $\# \{ i \mid 0 \le i < N \land a[i] = 1 \}$ and use the
  tick metric that assigns cost $n$ to the statement \code{tick(n)} and
  0 to all other operations.
  }
\label{fig:xmplinc}
\end{figure}
%
\iffull{
}{}
We can use the logic to prove a concrete bound for a given resource
metric, or a bound that is parametric in a set of resource metrics.
For example, it is possible to derive the following quantitative
triple for a binary-search function \code{bsearch}.% that holds for all
%stack metrics $M$.  
%
\begin{center}
${ \{ (Z = \log_2(h-\ell)) + Z{\cdot}M_\code{bsearch} \} }$
\code{bsearch(x,l,h)}
${ \{ Z{\cdot}M_\code{bsearch} \} }$
\end{center}
%
It holds for all stack metrics $M$.
A stack metric assigns cost $0$ to all evaluation steps except
function calls.  Before a call $f(x)$, we consume $M_f > 0$ resources
and after the call we return $M_f$ resources that can be used in
subsequent function calls.  So $M_f$ corresponds to the stack-frame
size of the function $f$.  In the pre- and postcondition, the logical
variable $Z$ is used to relate the size of the input with the
potential that is left after the function call.  The logical part $Z =
\log_2(h-\ell)$ of the precondition is $0$ if $Z =
\log_2(\state_0(h)-\state_0(\ell))$ in the initial state $\state_0$ and
$\infty$ otherwise.

@@

A program logic provides a principled foundation for statically
analyzing programs.  However, program logics need to be supported by
automatic methods to be useful in practice.  That is why we developed a
new automatic amortized analysis for deriving quantitative Hoare
triples for common C programs.

@@

This idea is best explained by example.  If we again use the
\emph{ticks} metric that assigns cost $n$ to the function call
\code{tick(n)}\iffull{ and cost $0$ to all other operations}{} then
the cost of the following example can be bounded by $|[x,y]|$. Note
that there are no restrictions on the signs of the input variables in
the examples.
\begin{lstlisting}[basicstyle=\tt\small]
while (x<y) { x=x+1; tick(1); } #\hfill \textnormal{(Example 1)}#
\end{lstlisting}
To derive this bound in our automatic amortized analysis, we start
with the initial potential $\Phi_0 = |[x,y]|$ (short for
$\Phi_0(\state) = |[\state(x),\state(y)]|$) which we also use as the
loop invariant.  For the loop body we then (like in Hoare logic) have
to derive a triple like
$\htriple{\Phi_0}{\code{x=x+1;tick(1)}}{\Phi_0}$.  We can
only do so if we utilize the fact that $x<y$ at the beginning of the
loop body.  % The reasoning then works as follows. We start with the
% potential $|[x,y]|$ and the fact that $|[x,y]| > 0$ before the
% assignment.  
If we denote the updated version of $x$ after the
assignment by $x'$ then the relation $|[x,y]| = |[x',y]| + 1$ between
the potential before and after the assignment \code{x=x+1} holds.
This means that we have the potential $|[x,y]| + 1$ before the
statement \code{tick(1)}.  Since \code{tick(1)} consumes one resource unit,
we end up with potential $|[x,y]|$ after the loop body and have
established the loop invariant again.

Note that our notion of a potential function is a generalization of the
concept of a ranking function.  A potential function can be used like
a ranking function if we use the tick metric and add the statement
\code{tick(1)} to every back edge of the program (loops and function
calls).  However, a potential function is more flexible.  For example,
we can use a potential function to prove that Example~2 does not
consume any resources in the tick metric.
\begin{lstlisting}[basicstyle=\tt\small]
  while (x<y) { tick(-1); x=x+1; tick(1); } #\hfill \textnormal{(Example 2)}#
\end{lstlisting}
Similarly we can prove that Example~3 can be bounded by $10|[x,y]|$.
In both cases, we reason exactly like in the first version of the
while loop to prove the bound.  Of course, such loops with different
tick annotations can be seamlessly combined in a larger program.
\begin{lstlisting}[basicstyle=\tt\small]
  while (x<y) { x=x+1; tick(10); } #\hfill \textnormal{(Example 3)}#
\end{lstlisting}
%
More formally, we develop (\pref{sec:AAA}) a judgement 
$$
\Gamma; Q \vdash S \dashv Q'; \Gamma'
$$
such that $\Gamma$ contains logical assertions like $x<y$ that we
collect along the conditional branches of the program and $Q$ is a
family of coefficents $Q = (q_{(a,b)})_{a,b \in \text{scope}}$ that is
indexed by the variables currently in scope.  The logical
context $\Gamma$ is simply a conjunction of inequalities between
linear combinations of variables.  It is used, for example, to determine
at the assignment if we can extract constant potential to pay for future
cost.  It is important to note that the reasoning about assertions in
$\Gamma$ is very basic.  We do not perform any fixpoint computations
and only derive trivial loop invariants about variables that are
unchanged in the loop body.

\pref{fig:ex1} shows a derivation of the bound
$\frac{T}{K}{\cdot}|[x,y]|$ on the number of ticks for a generalized
version of Example~1 in which increment $x$ be a constant $K>0$ and
that consumes $T>0$ resources in each iteration.  The reasoning is similar
to the one of Example~1 except that we obtain the potential
$K{\cdot}\frac{T}{K}$ after the assignment.  Note that the logical
assertions in $\Gamma$ are only used in rule for the assignment
\code{x=x+K}.  To the best of our knowledge, no other implemented tool
for C is currently capable of deriving a tight bound on the cost of
such a loop.  For $T=1$ (many systems focus on the number of loop
iterations without a cost model) and $K=10$,
KoAT~\cite{BrockschmidtEFFG14} computes the bound $x + y + 10$,
Rank~\cite{AliasDFG10} computes the bound $y-x-7$, and
LOOPUS~\cite{SinnZV14} computes the bound $y-(x+9)$.  Only
PUBS~\cite{AlbertAGPZ12} computes the tight bound $0.1(y-x)$ if we
translate the program into a term-rewriting system by hand.

To automate the reasoning, we first introduce an unknown rational
variable for each factor in the potential functions.  We then use our
inference rules (see \pref{sec:AAA}) to emit linear inequalities on
these variables that enforce that the potential annotations are sound.
For instance if $K = 1$ in \pref{fig:ex1} then we would have the
annotation $q_0 + q_{(x,y)} {\cdot} |[x,y]| + q_{(y,x)} {\cdot}
|[y,x]|$ before the assignment and $p_0 + p_{(x,y)} {\cdot} |[x,y]| +
p_{(y,x)} {\cdot} |[y,x]|$ after the assignment where $q_i$ and $p_i$
are unknown and must satisfy constraints like $p_{(x,y)} = q_{(x,y)}$,
$p_{(y,x)} = q_{(y,x)}$, and $p_0 = q_0 + q_{(x,y)} - q_{(y,x)}$.

\iffull{It might be surprising that we}{We} only track simple linear relations in
the logical context $\Gamma$.  It would of course be possible to keep
track of more sophisticated assertions.  However, this simple form is
sufficient for the examples we considered and we can efficiently
decide queries such as $\Gamma \implies x<y \; ?$ that we make in the
assignment rule.  Nevertheless, this logical part of our analysis
system is orthogonal to the quantitative part and can be easily
extended if necessary.

As mentioned, the automatic analysis can handle challenging example
programs without special tricks or techniques.  Examples
\emph{speed\_1} and \emph{speed\_2} that are taken from previous
work~\cite{GulwaniMC09} and demonstrate that our method can handle
\emph{tricky iteration patterns}.  The SPEED tool~\cite{GulwaniMC09}
derives the same bounds as our analysis but requires heuristics for
its counter instrumentation.  It can also be handled with inference of
\emph{disjunctive invariants}, but in the abstract interpretation
community, these invariants are known to be notoriously difficult to
handle.
%
In example \emph{speed\_1} we have one loop that first increments
variable $y$ up to $m$ and then increments variable $x$ up to $n$.  We
derive the tight bound $|[x, n]| + |[y, m]|$.
%
Example \emph{speed\_2} is more tricky and we found it hard to
manually find a bound.  However, using potential transfer reasoning as
in amortized analysis, it is easy to prove the tight bound
$|[x, n]| + |[z, n]|$.

Example \emph{t08a} shows the ability of the analysis to discover
interaction between \emph{sequenced loops} through size change of
variables.  We accurately track the size change of $y$ in the first
loop by transferring the potential $0.1$ form $|[y,z]|$ to $|[0,y]|$.
Furthermore, \emph{t08a} shows again that we do not handle the
constants $1$ or $0$ in any special way.  In all examples we could
replace $0$ and $1$ with other constants like in the second loop and
still derive a tight bound.  \iffull{The only information, that the analyzer
needs is $y \geq c$ before assigning $y = y - c$.}{}
%
Example \emph{t27} shows how amortization can be used to handle
\emph{interacting nested loops}.  In the outer loop we increment the
variable $n$ until $n = 0$.  In each of the $|[n,0]|$ iterations, we
increment the variable $y$ by $1000$.  Then we non-deterministically
(expressed by \code{*}) execute an inner loop that decrements $y$ by
$100$ until $y<100$.  The analysis discovers that only the first
execution of the inner loop depends on the initial value of $y$.  We
again derive tight constant factors.
%

As mentioned, the analysis also handles advanced control flow like
\code{break} and \code{return} statements, and mutual recursion.
\pref{fig:ex_rec} contains two mutually-recursive functions with their
automatically derived tick bounds.  The function \code{count\_down}
decrements its first argument $x$ until it reaches the second argument
$y$. It then recursively calls the function \code{count\_up}.  The
function \code{count\_up} is dual to \code{count\_down}.  Here, we
count up $y$ by $2$ and recursively call \code{count\_down}.  Our
analysis is the only available system that computes a tight
bound on this example.  \iffull{The analysis amounts to computing the meeting
point of two trains that approach each other with different speeds.}{}
% This is comparable to the
% determination of the point at which to trains meet if they approach
% each other with different speeds.

@@

Using the semantic validity of triples we define the validity
of a function context $\Delta$, written $\validC n \Delta$, as
\begin{align*}
  &\forall f .\, \Delta(f) =
    \forall z \, \vec v \, v .(P_f \, z \, \vec v, Q_f \, z \, v)
    \implies \\
  &\qquad \forall z \, \vec v .\,
  \valid n \bot {Q_f \, z} {P_f \, z \, \vec v} {S_f} \bot ,
\end{align*}
where $S_f$ is the body of the function $f$. \iffull{A full
judgement that mentions a non-empty function context
$\Delta$ is in fact a guarded statement: it makes
assumptions on some functions' behavior.}{}  The predicate
$\validC n \Delta$ gives the precise meaning of the
assumptions made.  It is also step-indexed to prove the
soundness of the {\sc L:Extend} rule by induction.
%
We are now able to state the soundness of the quantitative logic.
%
\begin{theorem}[Soundness of the logic]
  If $\Delta; B; R \vdash \htriple P S Q$ is derivable then
  $
    \forall n .\, \validC n \Delta
      \implies \valid {n {+} 1} B R P S Q.
  $
\end{theorem}
%
\noindent
The difference $\delta = 1$ between the index in the triple
validity and the one in context validity arises from
the soundness proofs of {\sc L:Call} and {\sc L:Extend}.  For
{\sc L:Call}, the language semantics makes one step and
proceeds with the function body, so we must have
$\delta \le 1$ to use the assumptions in $\Delta$.
For {\sc L:Extend}, we have to show that $\Delta \cup \Delta'$
is a valid context for $n$ steps.  The induction hypothesis
in that case says that if $\Delta \cup \Delta'$ is valid
for $m$ steps, $\Delta'$ is valid for $m+\delta$ steps.
So if we want to solve this goal by induction, it is
necessary that $\delta \ge 1$.  These two contraints force
$\delta$ to be exactly one in the theorem statement.

@@

% \paragraph{Program conditions.}
% The first pass in our OCaml implementation generates the
% logical annotations $\Gamma$ around all program statements.
% These annotations are used later on by the quantitative
% analysis to generate contraints.
% %
% The first design choice we had to make is to give a fixed form
% to program conditions.  Because of their generality and because
% they are closed under negation, we decided to consider
% inequalities over linear combinations as program conditions.
% \begin{mathpar}
% C := L < L \mid L > L \mid L \le L \mid L \ge L
% \and
% L := x \mid k \mid k * L \mid L + L \mid L - L
% \end{mathpar}
% A Condition is represented internally as a pair of one constant
% $k \in \mathbb N$ and one map $m : V \rightarrow \mathbb N$.
% Its intended semantics is
% $$
% \llbracket m \rrbracket(H) = k + \sum_v m_v \cdot H(v) \le 0.
% $$
% Note that conditions of this shape are closed by negation, as showed
% by the following equivalence
% $$
% \neg (k + \sum_v m_v \cdot H(v) \le 0)
% \Leftrightarrow (1-k) + \sum_v -m_v \cdot H(v) \le 0
% $$
% We can also note that all the ``questions'' asked to the logic
% assertions during a proof derivation can be described using
% linear combinations.  Conveniently enough, this language of
% assertions is exactly Presburger's arithmetic.  This means that
% we have a sound and complete decision procedure to answer
% all the questions asked during the analysis.

@@

% \subsection{Linear Contraints Generation}

>>>>

The idea that drives the design of our framework is amortized
analysis~\cite{Tarjan-amort}.  Assume that a program $S$ executes on a
starting state $\state$ and consumes $n$ resource units of some
user-defined quantity.  We denote that by writing $(S, \state)
\Downarrow_n \state'$ where $\state'$ is the program state after the
execution.  The basic idea of amortized analysis is to define a
\emph{potential function} $\Phi$ that maps program states to non-negative
numbers and to show that $\Phi(\state) \geq n$ if $\state$ is a
program state such that $(S, \state) \Downarrow_n \state'$ to prove 
a bound on the resource usage.

@@

\caption{Example derivation using amortized reasoning.  We write
  $\#_1(a)$ for $\# \{ i \mid 0 \le i < N \land a[i] = 1 \}$ and use the
  tick metric that assigns cost $n$ to the statement \code{tick(n)} and
  0 to all other operations.
  }
\label{fig:xmplinc}
\end{figure}
%
\iffull{
}{}
We can then use the logic to prove a concrete bound for a given resource
metric, or a bound that is parametric in a set of resource metrics.
For example, we can derive the following quantitative
triple for a binary-search function \code{bsearch} that holds for all
stack metrics $M$.  
%
\begin{center}
${ \{ (Z = \log_2(h-\ell)) + Z{\cdot}M_\code{bsearch} \} }$
\code{bsearch(x,l,h)}
${ \{ Z{\cdot}M_\code{bsearch} \} }$
\end{center}
%
A stack metric assigns cost $0$ to all evaluation steps except
function calls.  Before a call $f(x)$, we consume $M_f > 0$ resources
and after the call we return $M_f$ resources that can be used in
subsequent function calls.  So $M_f$ corresponds to the stack-frame
size of the function $f$.  In the pre- and postcondition, the logical
variable $Z$ is used to relate the size of the input with the
potential that is left after the function call.  The logical part $Z =
\log_2(h-\ell)$ of the precondition is $0$ if $Z =
\log_2(\state_0(h)-\state_0(\ell))$ in the initial state $\state_0$ and
$\infty$ otherwise.

@@

A program logic provides a principled foundation for statically
analyzing programs.  However, program logics need to be supported by
automatic methods to be useful in practice.  % That is why we developed a
% new automatic amortized analysis for deriving quantitative Hoare
% triples for common C programs.

@@

This idea is best explained by example.  If we again use the
\emph{ticks} metric that assigns cost $n$ to the function call
\code{tick(n)}\iffull{ and cost $0$ to all other operations}{} then
the cost of the following example can be bounded by $|[x,y]|$.\footnote{Note
that there are no restrictions on the signs of the input variables in
the examples.}
\begin{lstlisting}[basicstyle=\tt\small]
while (x<y) { x=x+1; tick(1); } #\hfill \textnormal{(Example 1)}#
\end{lstlisting}
To derive this bound in our automatic amortized analysis, we start
with the initial potential $\Phi_0 = |[x,y]|$ (short for
$\Phi_0(\state) = |[\state(x),\state(y)]|$) which we also use as the
loop invariant.  For the loop body we then (like in Hoare logic) have
to derive a triple like
$\htriple{\Phi_0}{\code{x=x+1;tick(1)}}{\Phi_0}$.  We can
only do so if we utilize the fact that $x<y$ at the beginning of the
loop body.  % The reasoning then works as follows. We start with the
% potential $|[x,y]|$ and the fact that $|[x,y]| > 0$ before the
% assignment.  
If we denote the updated version of $x$ after the
assignment by $x'$ then the relation $|[x,y]| = |[x',y]| + 1$ between
the potential before and after the assignment \code{x=x+1} holds.
This means that we have the potential $|[x,y]| + 1$ before the
statement \code{tick(1)}.  Since \code{tick(1)} consumes one resource unit,
we end up with potential $|[x,y]|$ after the loop body and have
established the loop invariant again.

Note that our notion of a potential function is a generalization of the
concept of a ranking function.  A potential function can be used like
a ranking function if we use the tick metric and add the statement
\code{tick(1)} to every back edge of the program (loops and function
calls).  However, a potential function is more flexible.  For example,
we can use a potential function to prove that Example~2 does not
consume any resources in the tick metric.
\begin{lstlisting}[basicstyle=\tt\small]
  while (x<y) { tick(-1); x=x+1; tick(1); } #\hfill \textnormal{(Example 2)}#
\end{lstlisting}
Similarly we can prove that Example~3 can be bounded by $10|[x,y]|$.
In both cases, we reason exactly like in the first version of the
while loop to prove the bound.  Of course, such loops with different
tick annotations can be seamlessly combined in a larger program.
\begin{lstlisting}[basicstyle=\tt\small]
  while (x<y) { x=x+1; tick(10); } #\hfill \textnormal{(Example 3)}#
\end{lstlisting}
%
More formally, we develop (\pref{sec:AAA}) a judgement 
$$
\Gamma; Q \vdash S \dashv Q'; \Gamma'
$$
such that $\Gamma$ contains logical assertions like $x<y$ that we
collect along the conditional branches of the program and $Q$ is a
family of coefficents $Q = (q_{(a,b)})_{a,b \in \text{scope}}$ that is
indexed by the variables currently in scope.  The logical
context $\Gamma$ is simply a conjunction of inequalities between
linear combinations of variables.  It is used, for example, to determine
at variable assignments if we can extract constant potential to pay for future
cost.  It is important to note that the reasoning about assertions in
$\Gamma$ is very basic.  We do not perform any fixpoint computations
and only derive trivial loop invariants about variables that are
unchanged in the loop body.

\pref{fig:ex1} shows a derivation of the bound
$\frac{T}{K}{\cdot}|[x,y]|$ on the number of ticks for a generalized
version of Example~1 in which we increment $x$ by a constant $K>0$ and
consume $T>0$ resources in each iteration.  The reasoning is similar
to the one of Example~1 except that we obtain the potential
$K{\cdot}\frac{T}{K}$ after the assignment.  Note that the logical
assertions in $\Gamma$ are only used in rule for the assignment
\code{x=x+K}.  To the best of our knowledge, no other implemented tool
for C is currently capable of deriving a tight bound on the cost of
such a loop.  For $T=1$ (many systems focus on the number of loop
iterations without a cost model) and $K=10$,
KoAT~\cite{BrockschmidtEFFG14} computes the bound $x + y + 10$,
Rank~\cite{AliasDFG10} computes the bound $y-x-7$, and
LOOPUS~\cite{SinnZV14} computes the bound $y-(x+9)$.  Only
PUBS~\cite{AlbertAGPZ12} computes the tight bound $0.1(y-x)$ if we
translate the program into a term-rewriting system by hand.

To automate the reasoning, we first introduce an unknown rational
variable for each factor in the potential functions.  We then use our
inference rules (see \pref{sec:AAA}) to emit linear constraints on
these variables that enforce that variable assignments that respect
the constraints correspond to sound potential annotations.  For
instance if $K = 1$ in \pref{fig:ex1} then we would have the
annotation $q_0 + q_{(x,y)} {\cdot} |[x,y]| + q_{(y,x)} {\cdot}
|[y,x]|$ before the assignment and $p_0 + p_{(x,y)} {\cdot} |[x,y]| +
p_{(y,x)} {\cdot} |[y,x]|$ after the assignment, where $q_i$ and $p_i$
are unknown and must satisfy constraints like $p_{(x,y)} = q_{(x,y)}$,
$p_{(y,x)} = q_{(y,x)}$, and $p_0 = q_0 + q_{(x,y)} - q_{(y,x)}$.

\iffull{It might be surprising that we}{We} only track simple linear relations in
the logical context $\Gamma$.  While it would of course be possible to keep
track of more sophisticated assertions, this simple form is
sufficient for the examples we considered and we can efficiently
decide queries such as $\Gamma \implies x<y \; ?$ that we make in the
assignment rule.  Nevertheless, this logical part of our analysis
system is orthogonal to the quantitative part and can be easily
extended if necessary.

As mentioned earlier, the automatic analysis can handle challenging example
programs without special tricks or techniques.  Examples
\emph{speed\_1} and \emph{speed\_2}, that are taken from previous
work~\cite{GulwaniMC09},  demonstrate that our method can handle
\emph{tricky iteration patterns}.  The SPEED tool~\cite{GulwaniMC09}
derives the same bounds as our analysis but requires heuristics for
its counter instrumentation.  It can also be handled with inference of
\emph{disjunctive invariants}, but in the abstract interpretation
community, these invariants are known to be notoriously difficult to
handle.
%
In example \emph{speed\_1} we have one loop that first increments
variable $y$ up to $m$ and then increments variable $x$ up to $n$.  We
derive the tight bound $|[x, n]| + |[y, m]|$.
%
Example \emph{speed\_2} is even trickier, and we found it hard to
find a bound manually.  However, using potential transfer reasoning as
in amortized analysis, it is easy to prove the tight bound
$|[x, n]| + |[z, n]|$.

Example \emph{t08a} shows the ability of the analysis to discover
interaction between \emph{sequenced loops} through size change of
variables.  We accurately track the size change of $y$ in the first
loop by transferring the potential $0.1$ from $|[y,z]|$ to $|[0,y]|$.
Furthermore, \emph{t08a} shows again that we do not handle the
constants $1$ or $0$ in any special way.  In all examples we could
replace $0$ and $1$ with other constants like in the second loop and
still derive a tight bound.  \iffull{The only information, that the analyzer
needs is $y \geq c$ before assigning $y = y - c$.}{}
%
Example \emph{t27} shows how amortization can be used to handle
\emph{interacting nested loops}.  In the outer loop we increment the
variable $n$ until $n = 0$.  In each of the $|[n,0]|$ iterations, we
increment the variable $y$ by $1000$.  Then we non-deterministically
(expressed by \code{*}) execute an inner loop that decrements $y$ by
$100$ until $y<100$.  The analysis discovers that only the first
execution of the inner loop depends on the initial value of $y$.  We
again derive tight constant factors.
%

As mentioned, the analysis also handles advanced control flow like
\code{break} and \code{return} statements, and mutual recursion.
\pref{fig:ex_rec} contains two mutually-recursive functions with their
automatically derived tick bounds.  The function \code{count\_down}
decrements its first argument $x$ until it reaches the second argument
$y$. It then recursively calls the function \code{count\_up},
which is dual to \code{count\_down}.  Here, we
count up $y$ by $2$ and recursively call \code{count\_down}.  Our
analysis is the only available system that computes a tight
bound on this example.  \iffull{The analysis amounts to computing the meeting
point of two trains that approach each other with different speeds.}{}
% This is comparable to the
% determination of the point at which to trains meet if they approach
% each other with different speeds.

@@

Using the semantic validity of triples we define the validity
of a function context $\Delta$, written $\validC n \Delta$, as
\begin{align*}
  &\forall f .\, \Delta(f) =
    \forall z \, \vec v \, v .(P_f \, z \, \vec v, Q_f \, z \, v)
    \implies \\
  &\qquad \forall z \, \vec v .\,
  \valid n \bot {Q_f \, z} {P_f \, z \, \vec v} {S_f} \bot ,
\end{align*}
where $S_f$ is the body of the function $f$. \iffull{A full
judgement that mentions a non-empty function context
$\Delta$ is in fact a guarded statement: it makes
assumptions on some functions' behavior.}{}  The predicate
$\validC n \Delta$ gives the precise meaning of the
assumptions made.  It is also step-indexed to prove the
soundness of the {\sc L:Extend} rule by induction.
%
We are now able to state the soundness of the quantitative logic.
%
\begin{theorem}[Soundness of the logic]
  If $\Delta; B; R \vdash \htriple P S Q$ is derivable then
  $
    \forall n .\, \validC n \Delta
      \implies \valid {n {+} 1} B R P S Q.
  $
\end{theorem}
%
\noindent
The difference $\delta = 1$ between the index in the triple
validity and the one in context validity arises from
the soundness proofs of {\sc L:Call} and {\sc L:Extend}.  For
{\sc L:Call}, the language semantics makes one step and
proceeds with the function body, so we must have
$\delta \le 1$ to use the assumptions in $\Delta$.
For {\sc L:Extend}, we have to show that $\Delta \cup \Delta'$
is a valid context for $n$ steps.  The induction hypothesis
in that case says that if $\Delta \cup \Delta'$ is valid
for $m$ steps, $\Delta'$ is valid for $m+\delta$ steps.
So if we want to solve this goal by induction, it is
necessary that $\delta \ge 1$.  These two constraints force
$\delta$ to be exactly one in the theorem statement.

@@

% \paragraph{Program conditions.}
% The first pass in our OCaml implementation generates the
% logical annotations $\Gamma$ around all program statements.
% These annotations are used later on by the quantitative
% analysis to generate constraints.
% %
% The first design choice we had to make is to give a fixed form
% to program conditions.  Because of their generality and because
% they are closed under negation, we decided to consider
% inequalities over linear combinations as program conditions.
% \begin{mathpar}
% C := L < L \mid L > L \mid L \le L \mid L \ge L
% \and
% L := x \mid k \mid k * L \mid L + L \mid L - L
% \end{mathpar}
% A Condition is represented internally as a pair of one constant
% $k \in \mathbb N$ and one map $m : V \rightarrow \mathbb N$.
% Its intended semantics is
% $$
% \llbracket m \rrbracket(H) = k + \sum_v m_v \cdot H(v) \le 0.
% $$
% Note that conditions of this shape are closed by negation, as showed
% by the following equivalence
% $$
% \neg (k + \sum_v m_v \cdot H(v) \le 0)
% \Leftrightarrow (1-k) + \sum_v -m_v \cdot H(v) \le 0
% $$
% We can also note that all the ``questions'' asked to the logic
% assertions during a proof derivation can be described using
% linear combinations.  Conveniently enough, this language of
% assertions is exactly Presburger's arithmetic.  This means that
% we have a sound and complete decision procedure to answer
% all the questions asked during the analysis.

@@

% \subsection{Linear Constraints Generation}
